Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2025-01-27 ms-3/ms.tex
Update ms.tex

\begin{abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 United States (US) counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

@@

An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see Supplemental Information). 
For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.

@@

\section*{Methods}
By definition incidence volatility is fast relative to broadscale epidemic dynamics. 
Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}. To allow for diverse shapes in the broadscale epidemic dynamics, these are fitted within a moving window
\begin{equation}
  \log\left(\dfrac{\mu_t}{N}\right)  = \sum_{j=1}^J \beta_j^{(t)} h_j(t)
\end{equation}
where $N$ represents population size, $h_j(t)$ are basis functions, $J$ is the degrees of freedom for the splines, and $\beta_j^{(t)}$ are fitted parameters for a symmetrical window of half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$. 
The degrees of freedom to be used for the splines, and the width of the moving window will depend on the application.
Explaination of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.

>>>>

\begin{Abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 United States (US) counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

@@

An interpretation of the dispersion parameter for a time series model of counts is that events are $1 + \theta^{-1}$ times as ``crowded'' in time relative to a Poisson process with the same mean~\citep{lloyd1967mean} (see derivation in S1 Text). 
For example, $\theta = 1$ corresponds to a situation where the average number of infections in the same time step as a randomly selected case will exceed the Poisson expectation by a factor of two. 
In a simple example relevant to surge capacity in healthcare systems, $\theta = 1$ implies that a random infectious individual visiting the emergency department at a hospital would find it on average to be twice as crowded with other infectious individuals (infected by the same pathogen) than expected for a Poisson process with the same incidence rate.

@@

\section*{Methods}
By definition incidence volatility is fast relative to broadscale epidemic dynamics. 
Consequently, in order to estimate incidence volatility we first model incidence at broad spatiotemporal scales using natural splines~\citep{perperoglou2019review}. To allow for diverse shapes in the broadscale epidemic dynamics, these are fitted within a moving window
\begin{equation}
  \log\left(\dfrac{\mu_t}{N}\right)  = \sum_{j=1}^J \beta_j^{(t)} h_j(t)
\end{equation}
where $N$ represents population size, $h_j(t)$ are basis functions, $J$ is the degrees of freedom for the splines, and $\beta_j^{(t)}$ are fitted parameters for a symmetrical window of half-width $\Delta$, centered at $t$, i.e., extending from $t-\Delta$ to $t + \Delta$. 
The degrees of freedom to be used for the splines, and the width of the moving window will depend on the application.
Explanation of the specific choices we used $J$ and $\Delta$ for our application to COVID-19 cases in US counties is described below.
